%!TEX root = INRL20160_Syllabus_1920.tex

\section*{Course reading}

\subsection*{Required readings:}

The following texts shall be used extensively throughout the course, so it is recommended that they are purchased:

\begin{itemize}
	\item Cini, M. and Borragán, N.P.S. 2019. European Union Politics. Oxford University Press.
	\item Hix, S. and H\o yland, B. 2011. The political system of the European Union. Palgrave Macmillan. 3rd Ed.
\end{itemize}

\subsection*{Recommended Readings}

The following books provide a general overview of the topics that will be covered in this module. 

\begin{itemize}
	\item Wallace, H., Pollack, M.A. and Young, A.R. eds. 2015. Policy-making in the European Union. Oxford University Press, USA.
	\item Bomberg, E., Peterson, J. and Corbett, R. 2012. The European Union: how does it work?. Oxford University Press.
	\item Bache, I., Bulmer, S., George, S. and Parker, O. 2014. Politics in the European Union. Oxford University Press, USA.
	\item Hix, S. 2013. What's Wrong with the Europe Union and How to Fix it. John Wiley \& Sons.
	\item Majone, G. 2009. Dilemmas of European integration: the ambiguities and pitfalls of integration by stealth. Oxford: Oxford University Press.
	\item Peterson, J. and Shackleton, M. 2012. The institutions of the European Union. Oxford University Press.
	\item Thomson, R. 2011. Resolving controversy in the European Union: legislative decision-making before and after enlargement. Cambridge University Press.
	\item Wiener, A. and Diez, T. 2009. European integration theory. Oxford: Oxford University Press.
\end{itemize}


In addition to these readings, students should keep up to date on current European affairs by reading daily newspapers, or one of the many websites devoted to EU politics. This reading is essential as it will allow you to keep up to date with current affairs in the EU and identify potential blog post topics. These websites include the following:

\begin{itemize}
	\item \url{http://www.euobserver.com}
	\item \url{http://europa.eu.int}
	\item \url{http://www.eupolitics.com}
	\item \url{http://www.ft.com}
	\item \url{http://blogs.lse.ac.uk/europpblog/}
	\item \url{http://blogs.lse.ac.uk/brexit/}
\end{itemize}

\section*{Brightspace}

Please make sure you have access to the module in Brightspace as soon as possible. It is the student's responsibility to make sure that they are signed up to the module correctly and they know how to submit coursework through the appropriate Brightspace assignment tab. If you have any issues with Brightspace contact the Brightspace support people to resolve the issue.

Furthermore, module materials such as this syllabus and announcements made outside lectures shall be on Brightspace. As such, Brightspace is an important communication tool for the module.

\subsection*{Detailed course programme}

\subsubsection*{Week 1}

\input{Topic_list/Intro_history}

\subsubsection*{Week 2 and Week 3}

\input{Topic_list/integration_theories}

%\subsubsection*{Week 3}

%\input{Topic_list/integration_theories}

\subsubsection*{Week 4}

\input{Topic_list/executive_politics}

\subsubsection*{Week 5}

\input{Topic_list/legislative}

\subsubsection*{Week 6}

\input{Topic_list/judicial}

\subsubsection*{Week 7}

\input{Topic_list/elections_pubopinion}

\subsubsection*{Week 8}

\input{Topic_list/brexit}

\subsubsection*{Week 9}

\input{Topic_list/igs}

\subsubsection*{Week 10}

\input{Topic_list/eurocrisis}

\subsubsection*{Week 11}

\input{Topic_list/demdeficit_transparency}

\subsubsection*{Week 12}

Revision and exam preparation
%\input{Topic_list/brexit}





