%!TEX root = INRL20160_Syllabus_1920.tex

\subsubsection*{Introduction and the history of European integration}

Study questions

\begin{itemize}
	\item What are the major challenges, threats, and opportunities facing the EU today? How do you think these issues will affect the EU and shape its future?
	\item What are the main milestones that the integration process in Europe went through? 
	\item How has the European integration project developed over time and how has the policy remit of the EU evolved?
\end{itemize}

\noindent Key reading

\begin{itemize}
	\item Hix, S. and H\o yland, B. 2011. The political system of the European Union. Palgrave Macmillan. Chapter 1.
	\item Cini, M. and Borragán, N.P.S. 2019. European Union Politics. Oxford University Press. Chapter 1-3.
\end{itemize}

\noindent Further reading

\begin{itemize}
	\item Urwin, D.W. 2014. The community of Europe: A history of European integration since 1945. Routledge.
	\item Dinan, D. 2010. Ever Closer Union: An Introduction to European Integration. London: Palgrave Macmillan. Chapters 1-6.
	\item Marks, G. 2011. Europe and Its Empires: From Rome to the European Union. Journal of Common Market Studies 50(1), pp.1-20.
	\item Nugent, N. 2010. The Government and Politics of the European Union. Basingstoke: Palgrave Macmillan, Part 1: Chapters 1-7.
	\item Bache, I. and George, S. 2011. Politics in the European Union. Oxford: Oxford University Press, Chapters 5-11.
	\item Diebold, W. 1959. The Schuman Plan. New York: Praeger.
	\item Dougan, M. 2008. The Treaty of Lisbon 2007: Winning Minds not Hearts. Common Market Law Review 45, pp.617–703.
	\item Gray, M. and Stubb, A. 2001. Keynote Article: The Treaty of Nice–Negotiating a Poisoned Chalice?. JCMS: Journal of Common Market Studies, 39(s1), pp.5-23.
	\item Magnette, P. and Nicolaidis, K. 2004. The European Convention: bargaining in the shadow of rhetoric. West European Politics, 27(3), pp.381-404.
	\item Majone, G. 2006. The Common Sense of European Integration. Journal of European Public Policy 13(5), pp.607-626.
	\item Piris, J-C. 2010. The Lisbon Treaty: a legal and political analysis. Cambridge: Cambridge University Press.
\end{itemize}